Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-04-10 TEBenchmarkPaper/article.tex
Move historical section into discussion
Added FIXMEs for things brought up in meeting

% Lay out the field of TE: motivation, tools, evaluation, problems...

@@

This has applications wherever we find use for formal statements: in proof
assistants and their libraries, in mathematics education and research, and in
the specification, verification, optimisation and testing of software.

Existing attempts at tackling this problem are difficult to compare, due
partially to the variety of approaches taken, but also because of the inherent
ambiguity of the task and the different goals emphasised by their designers and
their choice of evaluation method.

We attempt to solve this discrepancy, at least for the foreseeable future, by
defining a standard, unambiguous benchmarking approach with which to compare
the conjecture generation of MTE tools. Our contributions include:

@@

We introduce the conjecture generation problem in more detail, along with the
difficulty of comparing existing solutions, in $\S$\ref{sec:background}. In
$\S$\ref{exploration-versus-formation} we give a history of the MTE field, and
its close cousin Automated Theory Formation (ATF), describing a variety of
existing approaches and clarifying the diverse terminology found in the
literature. We explain our proposal for a more general benchmark in
$\S$\ref{sec:proposal} and demonstrate its application to existing MTE tools
in $\S$\ref{sec:application}. Issues facing our approach, and the field in
general, are discussed in $\S$\ref{sec:discussion} and concluding remarks
are given in $\S$\ref{sec:conclusion}.

@@

@@

@@

The task of automatically proving or disproving a \emph{given} conjecture has
been studied extensively in the field of Automated Theorem Proving (ATP). Less
attention has been paid to \emph{generating} such conjectures automatically,
either from a given set of definitions (the task we will focus on) or where the
definitions are also generated. The fields of Automated Theory Formation (ATF)
and Mathematical Theory Exploration (MTE) both tackle these generation tasks,
and we discuss their relationship in historical context in
$\S$\ref{exploration-versus-formation}.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

@@

\section{Theory Exploration versus Theory
  Formation}\label{exploration-versus-formation}

In this section we put the Mathematical Theory Exploration approach
into context both historically, and with respect to related approaches
such as in Automated Theory Formation tools.

\subsection{A Short History of Automated Mathematics}

\begin{quote}
...``in his subsequent design for an Analytical Engine Mr. Babbage has
shown that material machinery is capable, in theory at least, of
rivalling the labours of the most practised mathematicians in all
branches of their science.''~\cite[p. 498]{jevons}
\end{quote}

The first automated mathematical tool, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early tools include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
tools were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

We limit our discussion to applications in mathematics, although there are
similar approaches being applied to the task of Computational \emph{Scientific}
Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}.
Scientific knowledge relies on inductive reasoning and experimental testing,
which (in principle) are not so important to a mathematical theory. However,
these techniques \emph{do} play an important role in many mathematical theory
formation and exploration tools: in particular, inductive reasoning allows a
tool to conjecture statements which may be too difficult for it to deductively
prove; whilst experimental testing, e.g. by checking universally-quantified
statements against some specific values, is a simple way to quickly discard many
obvious falsehoods generated by powerful but imperfect algorithms (as opposed
to, say, limiting ourselves to verified theorem generators).

The phrase Mathematical Theory Exploration (MTE) is a recent term
which seems to have originated with Buchberger and colleagues (see,
for example,~\cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm~\cite{buchberger:04}.

%IsaScheme
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of zero
($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
following scheme to invent the concept of multiplication:

QuickSpec version 2 replaces this two-step process with a single, iterative
algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

We believe that ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches seem to be
that MTE was initiated by mathematicians as an attempt to support existing
mathematical research, perhaps with user interaction; while ATF is a product of
AI researchers trying to extend AI techniques, with full automation and often
applications to non-mathematical domains.

@@

\begin{enumerate}
\item The quality of each individual conjecture, which we will refer to as their
  ``interestingness''.
\item The quality of the \emph{set} of generated conjectures, which is important
  for criteria like novelty, as well as to assess consistency.
\item Resources used by the tool, to find the balance struck between output
  quality and time taken.
\end{enumerate}

@@

Tests were performed with a theory of natural numbers and a theory of lists (the
definitions shown in figure~\ref{figure:list_theory}), but it is difficult to
draw general conclusions from only two datapoints. These theories were also
quite small, the natural number theory
from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
definitions and 12 theorems.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should
  \emph{ideally} contain every ``interesting'' conjecture involving its included
  definitions. More \emph{realistically}, we should aim for each definition to
  appear in \emph{many} theorem statements; rather than each problem having
  unique definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

@@

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a commutativity theorem for a \texttt{plus} function, and an
associativity theorem for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
tool to either conjecture commutativity and associativity for \emph{both}
functions, or for \emph{neither} function, since they're logically equivalent.
Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

@@

\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

@@

\section{Application}
\label{sec:application}

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus~\cite{claessen2015tip}.

@@

\subsection{TIP}

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
light on this by testing whether mathematicians' experiences of abstract beauty
correlates with the same brain activity as experiences of sensory
beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
table~\ref{table:colton}, from computer scientists like Colton, are based
largely on ``intuition'', plausibility arguments about why a metric would be
important, and previous use of such metrics (in the case of
``surprisingingness'', from a single quote from a mathematician).
Opinions from mathematicians themselves include Gowers' suggestion that we can
identify features which are commonly associated with good
proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
book containing the most elegant proof of every theorem, brought to life
in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
developer).

All of these approaches rest upon the assumption that it even makes sense to
speak of ``the good'' in mathematics. However, empirical psychological studies
call into question such assumptions: for example, work by Inglis and colleagues
has shown that there is not even a single standard of \emph{validity} among
contemporary mathematicians~\cite{inglis2013mathematicians}.

@@

One notable weakness is that our methodology only makes sense in domains where
existing human knowledge surpasses that discovered by the machine. Any
\emph{truly novel} insights discovered during testing will not, by definition,
appear in any existing corpus, and we would in fact \emph{penalise} tools for
such output. We do not believe this to be a realistic problem for the time
being, although we would be delighted if proved wrong.

@@

@@

\begin{acknowledgements}
  We are grateful to the authors of the systems we have used in our experiments,
  especially for help in obtaining and configuring their tools. We would
  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
  their software to our purposes.
%  EPSRC
\end{acknowledgements}

>>>>

% Lay out the field of TE: motivation, tools, evaluation, problems...
% FIXME: Strengthen motivation:
%  - Automated Scientific Discovery (e.g. Dundee drug discovery "centaur" story)
%  - "Mixed initiative" (social machine?); maybe Kasparov
%  - Gowers calling for better AI in maths

@@

This has applications wherever we find use for formal statements: in proof
assistants and their libraries, in mathematics education and research, and in
the specification, verification, optimisation and testing of software. Similar
tasks are also studied in Computational \emph{Scientific}
Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling},
although that relies on inductive reasoning and experimental testing which (in
principle) are not so important for pure mathematics. Whilst we \emph{have}
found these methods to also be important for MTE tools (in particular for the
\emph{plausibility} of generated conjectures), we leave a more thorough
application of our results to scientific reasoning for future work.

Even limiting ourselves to mathematics applications, it is difficult to measure
and compare the success of approaches and tools. This is partially due to their
diversity, but also because of the inherent ambiguity of the task, the different
goals emphasised by their designers and the variety of evaluation methods
employed. We attempt to solve this discrepancy, at least for the foreseeable
future, by defining a standard, unambiguous benchmarking approach with which to
compare the conjecture generation of MTE tools. Our contributions include:

@@

We introduce the conjecture generation problem in more detail, along with the
difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We explain our proposal for a more general benchmark in
$\S$\ref{sec:proposal} and demonstrate its application to existing MTE tools
in $\S$\ref{sec:application}. Issues facing our approach, and the field in
general, are discussed in $\S$\ref{sec:discussion}; related work is surveyed in
$\S$\ref{sec:related-work}, including the diverse terminology found in the
literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.

@@

% FIXME: Perhaps mention examples like "plus(x, 0) = 0" as motivating examples
% of things which are "simple" but which proof assistants still don't do
% automatically (reference Moa's use of TE for background lemma generation)

@@

% FIXME: We jump from conjectures to theorems. Maybe to prevent confusion we
% should talk more generally about "interesting properties" of a formal system,
% and use that as motivation for conjecturing and proving?

@@

The task of automatically proving or disproving a \emph{given} conjecture has
been studied extensively in the field of Automated Theorem Proving (ATP). Less
attention has been paid to \emph{generating} such conjectures automatically,
either from a given set of definitions (the task we will focus on) or where the
definitions are also generated. The fields of Automated Theory Formation (ATF)
and Mathematical Theory Exploration (MTE) both tackle these generation tasks,
and we discuss their relationship in historical context in
$\S$\ref{related-work}.

@@

% FIXME: This is original work, so should be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

% FIXME: Give brief description and reference for each system in the text.
% We could include the more lengthy descriptions of IsaCoSy and QuickSpec here?

@@

@@

% FIXME: How are these referred to/considered in the paper?
\begin{enumerate}
\item The quality of each individual conjecture, which we will refer to as their
  ``interestingness''.
\item The quality of the \emph{set} of generated conjectures, which is important
  for criteria like novelty, as well as to assess consistency.
\item Resources used by the tool, to find the balance struck between output
  quality and time taken.
\end{enumerate}

@@

Tests were performed with a theory of natural numbers and a theory of lists (the
definitions shown in figure~\ref{figure:list_theory}), %FIXME: add page ref
but it is difficult to
draw general conclusions from only two datapoints. These theories were also
quite small, the natural number theory
from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
definitions and 12 theorems.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should
  \emph{ideally} contain every ``interesting'' conjecture involving its included
  definitions. More \emph{realistically}, we should aim for each definition to
  appear in \emph{many} theorem statements; rather than each problem having
  unique definitions. % FIXME: Reword: make clear that non-membership means
  % uninteresting, and hence the things we *don't* include are just as important
  % as the things we do
\item The problem set should be as large as possible, for robustness of the
  resulting statistics. % FIXME: Prioritising quantity over quality?
\end{itemize}

@@

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a commutativity theorem for a \texttt{plus} function, and an
associativity theorem for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
tool to either conjecture commutativity and associativity for \emph{both}
functions, or for \emph{neither} function, since they are logically equivalent.
Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

@@

% FIXME: Add section references to the braces
% FIXME: Maybe rethink "analysis" bit; maybe add summary and comparison
% references (although maybe not as full blown steps)
\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

@@

\section{Application}
\label{sec:application}
% FIXME: Maybe use (red?) crosses rather than red dots? More useful for black and white
We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus~\cite{claessen2015tip}.

@@

\subsection{TIP} % FIXME: Expand acronym

@@

% FIXME: The designers must decide on, not the system
Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
light on this by testing whether mathematicians' experiences of abstract beauty
correlates with the same brain activity as experiences of sensory
beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
table~\ref{table:colton}, from computer scientists like Colton, are based
largely on ``intuition'', plausibility arguments about why a metric would be
important, and previous use of such metrics (in the case of
``surprisingingness'', from a single quote from a mathematician).
Opinions from mathematicians themselves include Gowers' suggestion that we can
identify features which are commonly associated with good
proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
book containing the most elegant proof of every theorem, brought to life
in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
developer).

All of these approaches rest upon the assumption that it makes sense to speak of
``the good'' in mathematics. However, empirical psychological studies
call into question such assumptions: for example, work by Inglis and colleagues
has shown that there is not even a single standard of \emph{validity} among
contemporary mathematicians~\cite{inglis2013mathematicians}.

@@

One notable weakness is that our methodology only makes sense in domains where
existing human knowledge surpasses that discovered by the machine. Any
\emph{truly novel} insights discovered during testing will not, by definition,
appear in any existing corpus, and we would in fact \emph{penalise} tools for
such output. We do not believe this to be a realistic problem for the time
being, although we would be delighted if proved wrong. % FIXME: Is there a solution? Mention that this is common to most tests

@@

\section{Related Work}
\label{sec:related-work}

% FIXME: Reword into a short intro, e.g. "There is a long history of automating
% mathematics, going back to... A recurring theme is separation between the
% efforts of pure mathematicians and those wanting practical engineering
% applications....
The first automated mathematical tool, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early tools include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
tools were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

% FIXME: Cut this down a lot. Currently it's not even an ATF section, it's just
% an AM section. We should maybe remove its status as a section, and instead
% just remark on the influence of AM on the field

The phrase Mathematical Theory Exploration (MTE) is a recent term % FIXME: When?
which seems to have originated with Buchberger and colleagues (see, % FIXME: informal
for example,~\cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm~\cite{buchberger:04}.

% IsaScheme FIXME: Maybe strip down this description? We could use prose instead,
% like "schemes capture common patterns such as iteration, which can be
% instantiated with an addition function to produce multiplication, or with
% multiplication to produce exponentiation.
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of zero
($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
following scheme to invent the concept of multiplication:

% FIXME: CITE
QuickSpec version 2 replaces this two-step process with a single, iterative
algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

% FIXME: Condense
ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches seem to be
that MTE was initiated by mathematicians as an attempt to support existing
mathematical research, perhaps with user interaction; while ATF is a product of
AI researchers trying to extend AI techniques, with full automation and often
applications to non-mathematical domains.

@@

\begin{acknowledgements}
  We are grateful to the authors of the systems we have used in our experiments,
  especially for help in obtaining and configuring their tools. We would
  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
  their software to our purposes.
%  FIXME: EPSRC, Katya, Jianguo fundung, Alison funding
\end{acknowledgements}
